A Data-Driven Early Warning System for Mining Accident

### 2017-03-22 San Antonio/slides/presentation_slides.Rmd
updated markov by luo
semi-final slides

---
title: "A Data-Driven Early Warning System for Mining Accident"
shorttitle: "Mining Accident Data Analysis"
author: |
  | Yu Luo, Ashutosh Nanda, Shiva Rajgopal, Vinay Ramesh,
  | Venkat Venkatasubramanian, Zhizun Zhang, and Catherine Zhao
shortauthor: "Luo et al."
institute: |
  | Chemical Engineering, Computer Science, and Business School
  | Columbia University
shortinstitute: "Columbia University"
date: "3/27/2017"
output: 
  beamer_presentation:
    keep_tex: false
    incremental: false
    template: rmd_template.tex
    fig_height: 4.75
    toc: true
    slide_level: 2
    includes:
      in_header: rmd_header.tex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message=FALSE, fig.align = "center")
options(scipen = 1, digits = 2)
require(dplyr)
require(ggplot2)
knitr::opts_knit$set(root.dir = "~/Git/violation-data-analysis")
```

@@

<!-- https://aiche.confex.com/aiche/s17/webprogram/Paper480429.html -->
# Introduction: A Data Approach to Mine Safety

@@

@@

<!-- 
https://en.wikipedia.org/wiki/Upper_Big_Branch_Mine_disaster
https://arlweb.msha.gov/MEDIA/PRESS/2011/NR111206.asp 
-->
## Upper Big Branch Mine Disaster (2010)

@@

- Formed in	1977
- Agency of the United States Department of Labor
- Mission
    - Prevent death, illness, and injury from mining
    - Promote safe and healthful workplaces for U.S. miners
    - Develop and enforce safety and health rules
    - Provide technical, educational, and other types of assistance

@@

- Process MSHA safety data
- Understand the underlying causal relationships
- Develop early warning systems based on past behaviors
- Credit rating/score analogy
    - Predict default probability within 18 months
    - Accidents: defaults a month or a year prior to application
    - Violations: missed payments, late payments, etc.

@@

- Link: <https://enforcedata.dol.gov/views/data_catalogs.php>
- Updated daily or weekly
- Enforecement Data Catalog
    - MSHA Enforcement Data
    - OSHA Enforcement Data
    <!-- - Wage and Hour Compliance Action Data -->
    <!-- - OFCCP Compliance Evaluation and Complaint Investigation Data -->
    <!-- - EBSA Enforcement Data -->
    <!-- - Department of Justice -->
    - EPA
    - FDA

@@

- Each mine has a unique mine ID, e.g., Upper Big Branch (4608436)
- Rich details, e.g., selected attributes from the accidents table:

```{r msha accidents}
if(!exists("accidents")) {load("./San Antonio/data/secret_ingredients.RData")}
names(accidents[c(1, 14, 5:13, 19:20, 37:59)])
```

@@

- Missing data, typos
- Inactive mines are not labeled
- Most data are not numeric
- Relatively few severe accidents ($\sim0.5\%$)

@@

- Combine and summarize accidents/violations based on mines
- 664,128 rows, 10,377 unique mines
- From 2000 to 2015 in quarters
- Each row represents data for a unique combination of mine, year, and quarter
    - e.g., Upper Big Branch Mine in the second quarter of 2010
- Each row contains both current and past information

@@

- Number of days lost (`num.days.lost`), numer of days restricted (`num.days.restrict`), number of deaths (`num.death`), number of permenant disabilities (`num.dis`), and number of violations (`viol.quantity`)

@@

@@

@@

- Rising violation trends before disasters
- A disaster classifier based on historical data?
- Define a **severe** accident as one with death or permenant disability
- Unbalanced data

@@

- Biostatisticians and epidemiologists call it "conditional logistic regression" (`survival::clogit`)
- Suitable for **panel data** (e.g., our consolidated data)
- Model includes mine-specific but time-invariant variables
- Logistic regression (for every mine)

\[\Pr(Y=1|\mathbf{X})=F(\mathbf{X}) = \frac{1}{1+e^{-(\alpha+\boldsymbol\beta\mathbf{X})}}\]

- Logistic regression with fixed effects (for the $i$-th mine)

\[\Pr(Y=1|\mathbf{X},i)=F(\mathbf{x},i) = \frac{1}{1+e^{-(\alpha_i+\boldsymbol\beta\mathbf{x})}}\]

@@

<!-- 解释 confusion matrix和 specificity -->

@@

```{r glm}
if (!exists("glm.result")){load("./San Antonio/output/results_fixed_effects.RData")}
print(glm.performance$table)
print(glm.performance$byClass)
```

@@

- Fail to predict top 10 true positives

@@

- False positive predictions
<!-- - Top predicted positives (possibly false positives) -->

@@

```{r fe.div}
if (!exists("fe.div.result")){load("./San Antonio/output/results_fixed_effects.RData")}
print(fe.div.performance$table)
print(fe.div.performance$byClass)
```

@@

- Successfully predict all top 10 true positives

@@

- False positive predictions

@@

@@

@@

- Previously defined severe accidents and days lost $>$ 300

```{r fe.div2}
if (!exists("fe.div2.result")){load("./San Antonio/output/results_fixed_effects.RData")}
print(fe.div2.performance$table)
print(fe.div2.performance$byClass)
```

@@

- Successfully predict 9 out of top 10 true positives

@@

-

@@

## Markov Chain

@@

<!-- - Causal relationship between accidents and historical data -->
- Summary
    - Two deadliest mine accidents in the last decade: Upper Big Branch & Sago
    - MSHA 
- Application
    - "Credit score" for mine safety
    - Regulators, mines, stakeholders
- Future
    - Expand data: OSHA, EPA, etc.
    - Models and techniques: artificial neural networks (restricted boltzmann machine), text mining, etc.

>>>>

---
title: "A Data-Driven Early Warning System for Mining Accident"
shorttitle: "Mining Accident Data Analysis"
author: |
  | Yu Luo, Ashutosh Nanda, Shiva Rajgopal, Vinay Ramesh,
  | Venkat Venkatasubramanian, Zhizun Zhang, and Catherine Zhao
shortauthor: "Luo et al."
institute: |
  | Chemical Engineering, Computer Science, and Business School
  | Columbia University
shortinstitute: "Columbia University"
date: "3/27/2017"
output: 
  beamer_presentation:
    keep_tex: false
    incremental: false
    template: rmd_template.tex
    fig_height: 4.75
    # fig_width: 4
    toc: true
    slide_level: 2
    includes:
      in_header: rmd_header.tex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message=FALSE, fig.align = "center")
options(scipen = 1, digits = 2)
require(dplyr)
require(ggplot2)
require(markovchain)
knitr::opts_knit$set(root.dir = "~/Git/violation-data-analysis")
```

@@

<!-- https://aiche.confex.com/aiche/s17/webprogram/Paper480429.html -->
# Mine Safety: A Data-Driven Approach

@@

\placelogofalse

@@

<!-- https://en.wikipedia.org/wiki/Upper_Big_Branch_Mine_disaster -->
<!-- https://arlweb.msha.gov/MEDIA/PRESS/2011/NR111206.asp  -->
## Upper Big Branch Mine Disaster (2010)

@@

- Formed in	1977
- Agency of the United States Department of Labor
- Mission
    - Prevent death, illness, and injury from mining
    - Promote safe and healthful workplaces for U.S. miners
    - Develop and enforce safety and health rules
    - Provide technical, educational, and other types of assistance
- A constantly improving industry in terms of safety

@@

- Process MSHA safety data
- Understand the underlying causal relationships
- Develop early warning systems based on past behaviors
- Credit rating/score analogy
    - Predict default probability within 18 months
    - Accidents: defaults a month or a year prior to application
    - Violations: missed payments, late payments, etc.
- Can we develop a "credit score" for mine safety?

@@

- Link: <https://enforcedata.dol.gov/views/data_catalogs.php>
- Updated daily or weekly
- Publicly available 
    - Department of Labor: MSHA, OSHA, etc.
    - Other departments: EPA, FDA, DOJ, etc.

    <!-- - MSHA Enforcement Data -->
    <!-- - OSHA Enforcement Data -->
    <!-- - Wage and Hour Compliance Action Data -->
    <!-- - OFCCP Compliance Evaluation and Complaint Investigation Data -->
    <!-- - EBSA Enforcement Data -->
    <!-- - Department of Justice -->
    <!-- - EPA -->
    <!-- - FDA -->

@@

- Each mine has a unique mine ID, e.g., Upper Big Branch (4608436)
- Rich details: e.g., classification, description, and severity
- Selected attributes from the accidents table (omitting 42 attributes)

```{r msha accidents}
if(!exists("accidents")) {load("./San Antonio/data/secret_ingredients.RData")}
names(accidents[c(1, 14, 19:20, 5:13, 37, 39, 49:50)])
```

@@

- Missing data, human errors
- No information about inactive/nonoperating mines
- Most data are not numeric
- Lots of zeros, few severe accidents ($\sim0.5\%$)

@@

- Group and summarize accidents/violations by mines
- 664,128 rows, 10,377 unique mines
- From 2000 to 2015
- Each row represents data for a unique combination of mine, year, and quarter
    - e.g., Upper Big Branch Mine in the second quarter of 2010
- Each row contains both current and past information
    - i.e., current quarter, past quarter, past year, and past three years

@@

<!-- - Number of days lost (`num.days.lost`), numer of days restricted (`num.days.restrict`), number of deaths (`num.death`), number of permenant disabilities (`num.dis`), and number of violations (`viol.quantity`)  -->

- All 25 attributes of the consolidated data

@@

- Query the consolidated data on the deadliest accidents

@@

- Plot violation trends prior to Upper Big Branch and Sago disasters

@@

- Rising violation trends before disasters
- Develop a disaster classifier based on historical data
- Define a **severe** accident as one with death or permenant disability
<!-- - Unbalanced data -->

@@

- Biostatisticians and epidemiologists call it "conditional logistic regression" (`survival::clogit`)
- Suitable for **panel data** (e.g., longitudinal data, our consolidated data)
- Model includes mine-specific biases
- Logistic function (for every mine)

\[\Pr(Y=1|\mathbf{X}) = \frac{1}{1+e^{-(\alpha+\boldsymbol\beta\mathbf{X})}}\]

- Logistic function with fixed effects (for the $i$-th mine)

\[\Pr(Y=1|\mathbf{X},i) = \frac{1}{1+e^{-(\alpha_i+\boldsymbol\beta\mathbf{x})}}\]

@@

@@

```{r glm}
if (!exists("glm.result")){load("./San Antonio/output/results_fixed_effects.RData")}
print(glm.performance$table)
print(c(glm.performance$overall["Accuracy"], glm.performance$byClass[c("Sensitivity", "Specificity", "Precision", "F1")]))
```

- Accuracy = (TP + TN)/(P + N)
- Sensitivity/recall = TP/P
- Specificity = TN/N
- Precision = TP/(TP + FP)
- F1: harmonic mean of sensitivity and precision

@@

- Fails to predict top 10 deadliest disasters

@@

- List of false positive predictions based on predicted probability
<!-- - Top predicted positives (possibly false positives) -->

@@

```{r fe.div}
if (!exists("fe.div.result")){load("./San Antonio/output/results_fixed_effects.RData")}
print(fe.div.performance$table)
print(c(fe.div.performance$overall["Accuracy"], fe.div.performance$byClass[c("Sensitivity", "Specificity", "Precision", "F1")]))
```

@@

- Successfully predicts top 10 deadliest disasters

@@

- Still has a lot of false positive predictions
- List of false positive predictions based on predicted probability

@@

- What happened in the New Era Mine?

@@

- Among the worst mines by number of days lost due to accidents

```{r new-era-num-days-lost}
if(!exists("complete.active.quarters")){load("./San Antonio/output/result.RData")}
head(
  complete.active.quarters %>% filter(active) %>% arrange(desc(num.days.lost)) %>% select(
    mine.name, year, quarter, num.days.lost
  )
, 10)
```

- Rising violation trend from 2000 to 2005

@@

- Updated severe accident label
    - Previously defined criteria plus days lost $>$ 300
- Redo out-of-sample model

```{r fe.div2}
if (!exists("fe.div2.result")){load("./San Antonio/output/results_fixed_effects.RData")}
print(fe.div2.performance$table)
print(c(fe.div2.performance$overall["Accuracy"], fe.div2.performance$byClass[c("Sensitivity", "Specificity", "Precision", "F1")]))
```

- Worse true positive rate, improved F1 score

@@

- Successfully predicts 9 out of top 10 deadliest accidents

@@

- Accidents of the New Era mine are now true positives

@@

- Apply $k$-means clustering to consolidated data on all 20 features
- 3 clusters: `low`-risk, `mid`-risk, and `high`-risk
- Selected cluster centers (omitting 15 features)

```{r k-means}
if(!exists("clustering")){load("./San Antonio/output/results_markov.RData")}
sorted.cluster.centers <- clustering$centers[cluster.to.risk.label,c("num.days.lost", "num.days.restrict", "num.death", "num.dis", "viol.quantity")]
rownames(sorted.cluster.centers) <- risk.labels
print(sorted.cluster.centers)
```

- Cluster sizes

```{r k-means sizes}
if(!exists("clustering")){load("./San Antonio/output/results_markov.RData")}
sorted.cluster.size <- matrix(data = clustering$size[cluster.to.risk.label], nrow = 1, ncol = 3)
colnames(sorted.cluster.size) <- risk.labels
rownames(sorted.cluster.size) <- "size"
print(sorted.cluster.size)
```

<!-- https://cran.r-project.org/web/packages/markovchain/vignettes/an_introduction_to_markovchain_package.pdf -->
## Markov Chain

- Overall transition matrix

```{r markov-transition}
if(!exists("markov.mine")){load("./San Antonio/output/results_markov.RData")}
print(round(markov.mine[], 3))
```

- Steady-state distribution

```{r markov-steady-state}
if(!exists("markov.mine")){load("./San Antonio/output/results_markov.RData")}
print(round(steadyStates(markov.mine), 3))
```

@@

<!-- - Causal relationship between accidents and historical data -->
- Summary
    - Two deadliest mine accidents in the last decade: Upper Big Branch & Sago
    - Rich MSHA data that need clean-up
    - Supervised predictive model
    - Unsupervised clustering of risk
- Application
    - "Credit score" for mine safety
    - Regulators, mines, stakeholders
- Future
    - Improve model performance
    - Expand data: OSHA, EPA, etc.
    - Other techniques: artificial neural networks (restricted boltzmann machine), text mining, etc.
